Adaptive Waveform Inversion for Transmitted Wave Data

### 2024-11-07 repro/papers/awi/paper.tex
near final

Here, $u*f$ denotes the result of convolving each trace in $u$ by $f$. I
will also use the asterisk to denote the trace-by-trace convolution
of two data sets with the same set of source and receiver coordinates,
that is, the data set consisting of time convolutions of traces with the same source
and receiver coordinates. The meaning will be clear from context.

For such extended sources,
\[
  \bar{F}[\kappa,u*f] = u*F[\kappa]
\]
Introduce the notation $S[\kappa]$ for the operator of convolution
with $F[\kappa]$, that is,
\begin{equation}
  \label{eqn:sdef}
  S[\kappa]u = u*F[\kappa]
\end{equation}
$S$ represents an
extension of $F$, in the sense that with the choice
$u(\bx_r,t;\bx_s)=\delta(t)$, the output of $F$ is recovered:
\begin{equation}
  \label{eqn:sconsist}
  S[\kappa]\delta = F[\kappa].
\end{equation}

\cite{Warner:16} assume (implicitly) that there is always a unique
solution of the extended inversion problem, formulated in terms of
$S$: that is, that
for any $\kappa$ and any data $d$, there is an
adaptive filter $u[\kappa;d]$ for which
$S[\kappa]u[\kappa;d]= d$. If $d$ is noise-free, that is, $d=F[\kappa]$, then the consistency relation \ref{eqn:sconsist}
suggests that $u[\kappa;d](\bx_r,t;\bx_s) = \delta(t)$. This ideal adaptive filter is in the null
space of multiplication by $t$, that is, the operator
$u(\bx_r,t;\bx_s) \mapsto t u(\bx_r,t;\bx_s)$. This
observation motivates the definition of the AWI objective function
(\cite{Warner:16}, equation 4)::
\begin{equation}
  \label{eqn:awideforig}
  J_{\rm AWI}[\kappa;d] = \sum_{\bx_s,\bx_r} \left(\frac{\int dt |tu[\kappa;d](\bx_r,t;\bx_s)|^2}{\int dt |u[\kappa;d](\bx_r,t;\bx_s)|^2}\right) 
\end{equation}
in which $u[\kappa;d]$ is as defined above, that is, the solution of $S[\kappa]u[\kappa;d]
= d$. \cite{Warner:16} also formulate
AWI in terms of the inverse filter, that is, $u$ for which
$u*d = F[\kappa]$, and in fact recommend that definition. The present
discussion uses only the forward filter.
Adaptive waveform inversion consists in iterative minimization of
$J_{\rm AWI}$: choose an initial estimate $\kappa_0$ of the
model, then update $\kappa$ repeatedly using a gradient
descent method. \cite{Warner:16} describes the computation of $J_{\rm AWI}$
and its gradient in detail, and their use in a quasi-Newton
optimization algorithm.

The same observation motivates the definition of the {\em Matched
  Source Waveform Inversion} (MSWI) objective studied by
\cite{HuangSymes2015SEG,HuangSymes:Geo17}:
\begin{equation}
  \label{eqn:mswideforig}
  J_{\rm MSWI}[\kappa;d] = \sum_{\bx_s,\bx_r} \int dt |tu[\kappa;d](\bx_r,t;\bx_s)|^2
\end{equation}
That is, the two objectives differ only by the trace-by-trace
normalization featured in the definition of AWI (equation
\ref{eqn:awideforig}). Much of the analysis of $J_{\rm AWI}$ pertains to
$J_{\rm MSWI}$, as will be explained below.
The trace-by-trace normalization in the definition of $J_{\rm AWI}$
will prove quite consequential.

\section{Why it works}

It may not be obvious at first glance why AWI (local optimization of
$J_{\rm AWI}$) should be should be any more effective at estimating
$\kappa$ from data $d$ than is FWI (local
optimization of $J_{\rm FWI}$). Under the smoothness and ray geometry assumptions made above (equation
\ref{eqn:coeff}, conditions G1 - G4), an asymptotic relation
between AWI and travel time tomography emerges, justifying the notion
that AWI is more likely to succeed than is FWI.
%In addition all of the
%previously mentioned assumptions, I will assume that the data $d$ is
%{\em noise-free}, that is, $d = F[\kappa^*]$ for a target bulk modulus
%$\kappa^* \in M$.

\subsection{Tihonov Regularization}
Before establishing the relation with travel time tomography, several
defects in the definition \ref{eqn:awideforig} of $J_{\rm AWI}$ must be
remedied. First, the objective function is not well-defined for the
``ideal'' case, in which $d = F[\kappa]$ and $u[\kappa;d]=\delta$:
this adaptive filter is not square-integrable, so the denominator does
not make sense. Second, the equation $S[\kappa]u=d$ does not (in
general) have a solution $u \in L^2(X_{sr} \times \bR)$. Since
$S[\kappa]$ is a convolution operator, solving $S[\kappa]u=d$ for $u$
amounts to deconovolving $d$ by $F[\kappa]$..  Under the standing
assumption that $f \in C^{\infty}_0(\bR)$, $F[\kappa]$ does not have a continuous
(stable) convolution inverse, and $S[\kappa]$ is not surjective.

In fact, these two defects are really symptoms of the instability of
deconvolution for kernels with the properties of $F[\kappa]$.  From \ref{eqn:fwdgreen} and the
definition \ref{eqn:sdef} of  $S[\kappa]$,
\[
  S[\kappa]u = \rho( a[\kappa]\delta    +  (b[\kappa]H)) *\delta_{\tau[\kappa]}*\partial_t f *u,
\]
\begin{equation}
  \label{eqn:sgood}
  = \rho(a[\kappa]\delta_{\tau[\kappa]}*\partial_t f +
  (b[\kappa](0)\delta + (\partial_t b[\kappa])H)
  *\delta_{\tau[\kappa]}*f) * u.
\end{equation}
Since $\partial_t f$ is even, the formal adjoint $S[\kappa]^T$ is given by
\begin{equation}
  \label{eqn:sadjgood}
  S[\kappa]^Tu = \rho( a[\kappa]\delta    +  (\check{b}[\kappa]\check{H}))
  *\delta_{-\tau[\kappa]}*\partial_t f *u,
\end{equation}
in which $\check{u}(t) = u(-t)$.
The key properties of $S$ follow directly from Young's inequality and
the representation \ref{eqn:sgood}:
\begin{lemma}
  \label{thm:snorm}
  Under the assumptions G1-G4 leading to \ref{eqn:fwdgreen}, the
  expressions \ref{eqn:sgood} and \ref{eqn:sadjgood} define
  bounded operators on
  $\Pi_{X_{sr}} {\cal S}(\bR)$ and on
  $L^2(X_{sr} \times \bR)$, and there exists $C>0$ so
  that for any $\kappa \in M$,
  \begin{equation}
    \label{eqn:snorm}
    \|S[\kappa]\| \le C\|\partial_t f\|_{L^1(\bR)}.
  \end{equation}
\end{lemma}
\begin{proof} From the estimates H3, $b[\kappa](\bx_r,\cdot;\bx_s)H *
  \partial_tf \in {\cal S}(\bR)$ for all $(\bx_s,\bx_r) \in X_{sr}$,
  whence the first conclusion. Bounded extension to $L^2(X_{sr} \times
  \bR)$ follows from Young's inequality.
\end{proof}

The conventional remedy for instability in a deconvolution problem of
this type is so-called ``pre-whitening'', as
mentioned by \cite{Warner:16} ``Pre-whitening'' is a version of
Tihonov regularization. It  amounts to replacement of the equation
$S[\kappa]u=d$ by the regularized least squares problem: 
\begin{equation}
  \label{eqn:reg}
  u_{\sigma}[\kappa;d] = \mbox{argmin}_{u \in L^2(X_{sr} \times \bR)} (\|S[\kappa]u-d\|^2 + \sigma\|u\|^2)
\end{equation}
For future reference, examine the
implications of definition \ref{eqn:reg} for the units of the regularization weight $\sigma$,
which  is
necessarily dimensional. Since the data $d$ has dimensions of
pressure, so must $S[m]u = F[m]*u$. Of course $F[m]$ is also sampled
pressure, and convolution is a $t$ integration, so for $S[m]u$ to be a
pressure, $u$ must have dimensions of 1/time. Therefore
$\|S[m]u-d\|^2$ has dimensions of pressure$^2 \times $time, $\|u\|^2$ has
dimensions of 1/time, whence $\sigma$ must be dimensioned as
pressure$^2 \times$  time$^2$.

\begin{prop}
  \label{thm:unorm}
  Under the assumptions G1 - G4 justifying the
  representation \ref{eqn:fwdgreen}, the least squares problem
  \ref{eqn:reg} with $\sigma > 0$ has a unique solution
  $u_{\sigma}[\kappa;d] \in L^2(X_{sr} \times \bR)$ satisfying
  \begin{equation}
    \label{eqn:unorm}
    \|u_{\sigma}[\kappa;d]\| \le \frac{C}{\sigma}\|\partial_t f\|_{L^1(\bR)}\|d\|.
  \end{equation}
\end{prop}
\begin{proof}
For $\sigma > 0$, the Lax-Milgram Theorem (\cite{Yosida}, section
III.7) assures the existence of unique solution. 
\end{proof}

Define the modified
AWI objective $J_{\rm AWI,\sigma}$ by
replacing $u$ in the definition \ref{eqn:awideforig} with $u_{\sigma}$:
\begin{equation}
  \label{eqn:awidef}
  J_{\rm AWI,\sigma}[\kappa;d] = \sum_{\bx_s,\bx_r} \left(\frac{\int dt |tu_{\sigma}[\kappa;d](\bx_r,t;\bx_s)|^2}{\int dt |u_{\sigma}[\kappa;d](\bx_r,t;\bx_s)|^2}\right). 
\end{equation}
To make this definition a bit more concise, introduce the operator $T$ on
${\cal D}'(X_{sr} \times \bR)$ by
\[
  Tp(\bx_r,t;\bx_s) = tp(\bx_r,t;\bx_s).
\]
Then
\begin{equation}
  \label{eqn:awidefbis}
  J_{\rm AWI,\sigma}[\kappa;d] = \sum_{(\bx_s,\bx_r) \in X_{sr}}
  \frac{\|Tu_{\sigma}[\kappa;d](\bx_r,\cdot;\bx_s)\|^2}{\|u_{\sigma}[\kappa;d](\bx_r,\cdot;\bx_s)\|^2}.
\end{equation}
Here $\| \cdot \|$ is the norm in $L^2(\bR)$.

Define a regularized version of the MSWI objective function $J_{\rm
  MSWI}$ (definition \ref{eqn:mswideforig}) similarly:
\begin{equation}
  \label{eqn:mswidef}
  J_{\rm MSWI,\sigma}[\kappa;d] = \|Tu_{\sigma}[\kappa;d]\|^2.
\end{equation}

The definitions \ref{eqn:awidefbis} and \ref{eqn:mswidef} have introduced a new defect: there is no
guarantee that $Tu_{\sigma}[\kappa;d] \in L^2(X_{sr} \times\bR)$.

\begin{prop}
  \label{thm:tl2}
  Suppose that $d, Td \in L^2(X_{sr} \times \bR)$. Then
  $Tu_{\sigma}[\kappa;d]  \in L^2(X_{sr} \times \bR)$, and for each
  $(\bx_s,\bx_r) \in X_{sr}$,
  \[
    \|Tu_{\sigma}[\kappa;d](\bx_r,\cdot;\bx_s)\| \le
  \]
  \begin{equation}
    \label{eqn:tunorm}
    \frac{C}{\sigma}\left(\left(
        \frac{\|\partial_t f\|^2_{L^1(\bR)}}{\sigma}(\|f\|_{L^1(\bR)}+\|T\partial_tf\|_{L^1(\bR)})
      \right)\|d\| \\ 
    + \|\partial_t  f\|_{L^1(\bR)}(\|d (\bx_r,\cdot;\bx_s)\| +\|Td (\bx_r,\cdot;\bx_s)\|)\right).
  \end{equation}
\end{prop}

@@

\begin{prop}
  \label{thm:noisefree}
  Suppose that $\kappa \in M$. Then for $p \ge 1$, $F[\kappa], TF[\kappa]
  \in L^p(X_{sr} \times \bR)$, and there exist $C_0, C_1, C_2 >0$ so
  that for each $(\bx_s,\bx_r) \in X_{sr}$,
  \begin{eqnarray}
    \label{eqn:dnorms}
    C_1\|\partial_tf\|_{L^p(\bR)} - C_2\|f\|_{L^p(\bR)} \le \|F[\kappa](\bx_r,\cdot;\bx_s)\|_{L^p(\bR)} & \le & C_0\|\partial_tf\|_{L^p(\bR)}, \nonumber \\
    \|TF[\kappa](\bx_r,\cdot;\bx_s)\|_{L^p(\bR)} & \le & C_0\|\partial_tf\|_{L^p(\bR)}.
  \end{eqnarray}
\end{prop}

@@

Note that in the representation \ref{eqn:fwdgreen}, only the
derivative $\partial_tf$ appears. For convenience, introduce
\begin{equation}
  \label{eqn:wavelet}
  w = \rho \partial_t f
\end{equation}
Then
\begin{itemize}
\item[W1. ] $w \in C^{\infty}_0(\bR)$;;
\item[W2. ] $\int \, w = 0$.
\end{itemize}
I will also
refer to $w$ as ``the wavelet'', and view $F$ as parametrized by $w$.

@@

The modeling operator with wavelet $\wl$ is denoted $\Fl$. From the
decomposition \ref{eqn:fwdgreen}, $\Fl$ is the sum of two terms:
\begin{equation}
  \label{eqn:Flamdecomp}
  \Fl = \Fla + \Flb.
\end{equation}
where
\begin{equation}
  \label{eqn:Fl0}
  \Fla[\kappa]= a[\kappa]\delta_{\tau[\kappa]}*\wl
\end{equation}
and
\[
\Flb[\kappa] = (b[\kappa]H) *\delta_{\tau[\kappa]}*\wl.
\]
Write $b_0[\kappa](\bx_s,\bx_r) = b[\kappa](\bx_r,0;\bx_s)$. Then
$\Flb$ can be rearranged as
\begin{equation}
  \label{eqn:Fl1}
  =\rho (b_0[\kappa]\delta + (\partial_t b H)) * f_{\lambda} )*\delta_{\tau[\kappa]}.
\end{equation}
From the bounds \ref{eqn:best} and \ref{eqn:lambdanorms} and Young's
inequality, there exists $C \ge 0$ so that for all $\log \kappa \in M$, $p
\ge 1$,
\begin{equation}
  \label{eqn:Frembd}
 \|\Fla[\kappa]\|_{ L^p(X_{sr} \times \bR)} \le C\lambda^{\frac{1}{p}-\frac{1}{2}},\,\|\Flb[\kappa]\|_{ L^p(X_{sr} \times \bR)} \le C\lambda^{\frac{1}{p}+\frac{1}{2}}
\end{equation}

>>>>

\subsection{Tihonov Regularization}
Introduce the notation $S[\kappa]$ for the operator of convolution
with $F[\kappa]$, that is,
\begin{equation}
  \label{eqn:sdef}
  S[\kappa]u = u*F[\kappa]
\end{equation}
$S$ represents an
extension of $F$, in the sense that with the choice
$u(\bx_r,t;\bx_s)=\delta(t)$, the output of $F$ is recovered:
\begin{equation}
  \label{eqn:sconsist}
  S[\kappa]\delta = F[\kappa].
\end{equation}
The regularized least-squares problem \ref{eqn:reg} can be re-written
\begin{equation}
  \label{eqn:regs}
    u_{\sigma}[\kappa;d] = \mbox{argmin}_{u \in L^2(X_{sr} \times \bR)} (\|S[\kappa]u-d\|^2 + \sigma\|u\|^2)
  \end{equation}
\cite{Warner:16} present the definition of the AWI objective function
\ref{eqn:awidefmain} initially as if $\sigma=0$, in fact presuming a
zero-residual solution (that is, $S[\kappa]u=d$) but later note that
``pre-whitening'' is necessary for a well-behaved filter construction.
``Pre-whitening'' is a synonym for Tihonov regularization, that is,
the least-squares minimization \ref{eqn:regs} with $\sigma>0$.

Regularization has another benefit, in that it makes the objective
well-defined. Indeed, if the data is noise-free, that is,
$d=F[\kappa]=S[\kappa]\delta$, then $u_0[\kappa,d]=\delta$, but the
denominator in the definition \ref{eqn:awidefmain} is not defined.
Of course, the numerator vanishes, motivating the definition of AWI,
but the objective function is nonsense.

For future reference, examine the implications of definition
\ref{eqn:regs} for the units of the regularization weight $\sigma$,
which is necessarily dimensional. Since the data $d$ has dimensions of
pressure, so must $S[m]u = F[m]*u$. Of course $F[m]$ is also sampled
pressure, and convolution is a $t$ integration, so for $S[m]u$ to be a
pressure, $u$ must have dimensions of 1/time. Therefore
$\|S[m]u-d\|^2$ has dimensions of pressure$^2 \times $time, $\|u\|^2$
has dimensions of 1/time, whence $\sigma$ must be dimensioned as
pressure$^2 \times$ time$^2$.

From \ref{eqn:green} and the
definition \ref{eqn:sdef} of  $S[\kappa]$,
\[
  S[\kappa]u = \rho( a[\kappa]\delta    +  (b[\kappa]H)) *\delta_{\tau[\kappa]}*\partial_t f *u,
\]
\begin{equation}
  \label{eqn:sgood}
  = \rho(a[\kappa]\delta_{\tau[\kappa]}*\partial_t f +
  (b[\kappa](0)\delta + (\partial_t b[\kappa])H)
  *\delta_{\tau[\kappa]}*f) * u.
\end{equation}
The formal adjoint $S[\kappa]^T$ is given by
\begin{equation}
  \label{eqn:sadjgood}
  S[\kappa]^Tu = \rho( a[\kappa]\delta    -  (\check{b}[\kappa]\check{H}))
 *\delta_{-\tau[\kappa]}*\partial_t \check{f} *u,
\end{equation}
in which $\check{u}(t) = u(-t)$.
The key properties of $S$ follow directly from Young's inequality and
the representation \ref{eqn:sgood}:
\begin{lemma}
  \label{thm:snorm}
  Under the Exponential Decay, Single Arrival, and No Return assumptions, the
  expression \ref{eqn:sgood} define
  bounded operators on
  $\Pi_{X_{sr}} {\cal S}(\bR)$ and on
  $L^2(X_{sr} \times \bR)$, and there exists $C>0$ so
  that for any $\kappa \in M$,
  \begin{equation}
    \label{eqn:snorm}
    \|S[\kappa]\| \le C\|\partial_t f\|_{L^1(\bR)}.
  \end{equation}
\end{lemma}
\begin{proof} From the assumptions about $f$ and bounds
  \ref{eqn:best},  $b[\kappa](\bx_r,\cdot;\bx_s)H *
  \partial_tf \in {\cal S}(\bR)$ for all $(\bx_s,\bx_r) \in X_{sr}$,
  whence the first conclusion. Bounded extension to $L^2(X_{sr} \times
  \bR)$ follows from Young's inequality.
\end{proof}

\begin{prop}
  \label{thm:unorm}
  Under the assumptions of Lemma \ref{thm:snorm}, the least squares problem
  \ref{eqn:regs} with $\sigma > 0$ and $d\in L^2(X_{sr} \times \bR)$ has a unique solution
  $u_{\sigma}[\kappa;d] \in L^2(X_{sr} \times \bR)$ satisfying
  \begin{equation}
    \label{eqn:unorm}
    \|u_{\sigma}[\kappa;d]\| \le \frac{C}{\sigma}\|\partial_t f\|_{L^1(\bR)}\|d\|.
  \end{equation}
\end{prop}
\begin{proof}
For $\sigma > 0$, the Lax-Milgram Theorem (\cite{Yosida}, section
III.7) assures the existence of unique minimizer
$u_{\sigma}[\kappa;d]$ satisfying the  normal
equation
\begin{equation}
  \label{eqn:normal}
  (S[\kappa]^TS[\kappa] + \sigma I) u_{\sigma}[\kappa;d] = S[\kappa]^Td,
\end{equation}
and the bound follows from the inequality \ref{eqn:snorm}.
\end{proof}

\begin{prop}
  \label{thm:tl2}
  In addition to the Exponential Decay, Single Arrival, and No Return assumptions, uppose that $ Td \in L^2(X_{sr} \times \bR)$. Then
  $Tu_{\sigma}[\kappa;d]  \in L^2(X_{sr} \times \bR)$, and for each
  $(\bx_s,\bx_r) \in X_{sr}$,
  \[
    \|Tu_{\sigma}[\kappa;d](\bx_r,\cdot;\bx_s)\| \le
  \]
  \[
    \frac{C}{\sigma}\left(\left(
        \frac{\|\partial_t f\|^2_{L^1(\bR)}}{\sigma}(\|f\|_{L^1(\bR)}+\|T\partial_tf\|_{L^1(\bR)})
      \right)\|d\|\right.
  \]
  \begin{equation}
    \label{eqn:tunorm}
    \left.
    + \|\partial_t  f\|_{L^1(\bR)}(\|d (\bx_r,\cdot;\bx_s)\| +\|Td (\bx_r,\cdot;\bx_s)\|)\right).
  \end{equation}
\end{prop}

@@

\begin{prop}
  \label{thm:noisefree}
  Assume Exponential Decay, Single Arrival, and No Return. Then for
  $\log \kappa \in M$,r $p \ge 1$, $F[\kappa], TF[\kappa]
  \in L^p(X_{sr} \times \bR)$, and there exist $C_0, C_1, C_2 >0$ so
  that for each $(\bx_s,\bx_r) \in X_{sr}$,
  \begin{eqnarray}
    \label{eqn:dnorms}
    C_1\|\partial_tf\|_{L^p(\bR)} - C_2\|f\|_{L^p(\bR)} \le \|F[\kappa](\bx_r,\cdot;\bx_s)\|_{L^p(\bR)} & \le & C_0\|\partial_tf\|_{L^p(\bR)}, \nonumber \\
    \|TF[\kappa](\bx_r,\cdot;\bx_s)\|_{L^p(\bR)} & \le & C_0\|\partial_tf\|_{L^p(\bR)}.
  \end{eqnarray}
\end{prop}

@@

Note that in the first of the two representations in display \ref{eqn:sgood}, only the
derivative $\partial_tf$ appears. For convenience, introduce
\begin{equation}
  \label{eqn:wavelet}
  w = \rho \partial_t f
\end{equation}
Then  $w \in C^{\infty}_0(\bR)$ and $\int \, w = 0$. I will also
refer to $w$ as ``the wavelet'', and view $F$ as parametrized by $w$.

@@

The modeling operator with wavelet $\wl$ is denoted $\Fl$
(consistently with the prior definition using $\fl$).. From the
decomposition \ref{eqn:green}, $\Fl$ is the sum of two terms:
\begin{equation}
  \label{eqn:Flamdecomp}
  \Fl = \Fla + \Flb.
\end{equation}
where
\begin{equation}
  \label{eqn:Fl0}
  \Fla[\kappa]= a[\kappa]\delta_{\tau[\kappa]}*\wl
\end{equation}
and
\[
\Flb[\kappa] = (b[\kappa]H) *\delta_{\tau[\kappa]}*\wl.
\]
Write $b_0[\kappa](\bx_s,\bx_r) = b[\kappa](\bx_r,0;\bx_s)$. Then
$\Flb$ can be rearranged as
\begin{equation}
  \label{eqn:Fl1}
  =\rho (b_0[\kappa]\delta + (\partial_t b H)) * f_{\lambda} )*\delta_{\tau[\kappa]}.
\end{equation}
From the bounds \ref{eqn:best} and \ref{eqn:lambdanorms} and Young's
inequality, there exists $C \ge 0$ so that for all $\log \kappa \in M$, $p
\ge 1$,
\begin{equation}
  \label{eqn:Frembd}
 \|\Fla[\kappa]\|_{ L^p(X_{sr} \times \bR)} \le C\lambda^{\frac{1}{p}-\frac{1}{2}},\,\|\Flb[\kappa]\|_{ L^p(X_{sr} \times \bR)} \le C\lambda^{\frac{1}{p}+\frac{1}{2}}
\end{equation}
